\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath, amssymb, amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\PassOptionsToPackage{hyphens}{url}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue!50!black, urlcolor=blue!50!black, citecolor=blue!50!black,
  pdftitle={Differential transcendence of the Hahn-Exton q-cosine and the arithmetic of its zeros},
  pdfauthor={Vico Bonfioli}}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\C}{\mathbb{C}}
\DeclareMathOperator{\ord}{ord}
\DeclareMathOperator{\Cas}{Cas}

\title{Differential transcendence of the Hahn--Exton $q$-cosine\\ and the arithmetic of its zeros}
\author{Vico Bonfioli\thanks{Independent researcher. \texttt{vico@anvilstack.com}. Code and
verification data: \url{https://github.com/ElVec1o/pythagorean-reflection-sequence}.}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Let $G(q,z)={}_1\phi_1(0;q;q^2,z)=\sum_{k\ge0}(-1)^kq^{k(k-1)}z^k/(q;q)_{2k}$, the Hahn--Exton (third
Jackson) $q$-cosine. We prove that for every $0<q<1$ the difference Galois group over $\C(z)$ of the
second-order $q$-difference equation satisfied by $G$ contains $\mathrm{SL}_2$, and that the equation
admits no $\sigma\delta$-integrability relation; via the parametrized difference Galois theory of
Hardouin--Singer and Dreyfus--Hardouin--Roques it follows that $G(q,\cdot)$ is differentially
transcendental over $\C(z)$. We complement this function-level result with an exact evaluation of the
regularized product of the positive zeros, $\prod_{k\ge1}z_k(q)\,q^{2(k-1)}=(q;q^2)_\infty$, a modular
value. We then pose, and give effective evidence for, the arithmetic question that these results
frame: the transcendence of the zeros $z_k(q)$ at algebraic $q$. This question specializes to the
transcendence of a specific reflection-group growth constant $\beta_2$ (OEIS A396406), for which we
prove an unconditional effective non-rationality bound.
\end{abstract}

\section{Introduction}

The three Jackson $q$-analogues of the Bessel function are among the most-studied special functions of
$q$-series. The third, due to Hahn and Exton, has the $q$-cosine and $q$-sine specializations
\[
   G(q,z):={}_1\phi_1(0;q;q^2,z)=\sum_{k\ge0}\frac{(-1)^kq^{k(k-1)}}{(q;q)_{2k}}z^k,
\]
which is, up to an explicit prefactor, $J^{(3)}_{-1/2}(\sqrt z/q;q^2)$; it is entire of order $0$ in
$z$, and its positive zeros $0<z_1(q)<z_2(q)<\cdots$ are real and simple \cite{KS}. The analytic theory
of these zeros --- orthogonality, interlacing, asymptotics --- is classical. Their \emph{arithmetic}
is, to our knowledge, untouched: nothing is known about the transcendence of $z_k(q)$ at algebraic
$q$, the $q$-analogue of Siegel's theorem that the zeros of the Bessel function $J_\nu$ are
transcendental.

This paper establishes the function-theoretic groundwork for that arithmetic and states the resulting
transcendence problem precisely. Our results are of three kinds.

\emph{Galois theory (\S\ref{sec:galois}).} We compute enough of the difference Galois structure of the
defining equation to conclude differential transcendence.
\begin{theorem}\label{thm:main-transc}
For every $0<q<1$, the Hahn--Exton $q$-cosine $G(q,\cdot)$ satisfies no algebraic differential
equation over $\C(z)$.
\end{theorem}
The engine is an unconditional proof that the difference Galois group contains $\mathrm{SL}_2$
(Theorem~\ref{thm:sl2}), obtained by ruling out rank-$1$ submodules of every symmetric power through a
growth estimate on asymptotic-expansion coefficients, together with the absence of a
$\sigma\delta$-integrability relation (Proposition~\ref{prop:nonintegrable}).

\emph{A modular zero product (\S\ref{sec:zeroproduct}).} Writing $Q=q^2$,
\begin{theorem}\label{thm:zeroproduct-intro}
$\displaystyle\prod_{k\ge1}z_k(q)\,Q^{\,k-1}=(q;q^2)_\infty$, and $G(q,z)/\theta_Q(-z)\to1/(q;q)_\infty$
along the negative axis, where $\theta_Q(-z)=\sum_n(-1)^nq^{n(n-1)}z^n$.
\end{theorem}
The regularized product of the zeros is thus an explicit modular value --- transcendental at algebraic
$q$ by Nesterenko's theorem \cite{Nesterenko}.

\emph{Arithmetic of the zeros (\S\ref{sec:arith}).} These results frame a clean conjecture.
\begin{conjecture}[Z]\label{conj:Z}
For algebraic $q\in(0,1)$, every positive zero $z_k(q)$ of $G(q,\cdot)$ is transcendental.
\end{conjecture}
We explain why Conjecture~\ref{conj:Z} lies just past the reach of the $q$-analogue of the
Siegel--Shidlovskii method (the double-Pochhammer denominator $(q;q)_{2k}$ places it below the Siegel
margin of \cite{AMV}), and we record an unconditional partial result for the specific zero of
number-theoretic interest.

The motivating instance is the growth constant $\beta_2=1.4916177871\ldots$ of the geodesic growth
series of a planar right-triangle reflection group (OEIS A396406), which equals $1/\sqrt{q^\ast}$ where
$q^\ast$ is the least positive root of $S(q):=G(q,2q(1-q))$; equivalently $z^\ast:=2q^\ast(1-q^\ast)=
z_1(q^\ast)$ is the smallest positive zero of $G(q^\ast,\cdot)$. Conjecture~\ref{conj:Z} implies
$\beta_2$ transcendental. We prove unconditionally (Proposition~\ref{prop:effective}) that $q^\ast$ has
no rational representation of denominator $\le10^{191}$.

All numerical claims are reproducible from the repository above; each is annotated with the script that
verifies it and the precision attained.

\section{The $q$-difference module}\label{sec:prelim}

Throughout $Q=q^2$, $\sigma f(z)=f(Qz)$, and $a(z)=q+1-qz$. From the ${}_1\phi_1$ recurrence, $G$
satisfies
\begin{equation}\label{eq:Leq}
   q\,G(z)-a(z)\,G(Qz)+G(Q^2z)=0,\qquad L:=q-a\sigma+\sigma^2,
\end{equation}
a second-order $q$-difference equation with companion matrix $A(z)=\bigl(\begin{smallmatrix}0&1\\-q&
a\end{smallmatrix}\bigr)$, $\det A=q$, regular (Poincar\'e-type) at $z=0$ and irregular at $z=\infty$
(Newton slopes $\{0,1\}$). A second solution is $\widetilde G(z)=z^{1/2}H(z)$ with
$H(z)=\sum_{k\ge0}(-1)^kq^{k^2}(1-q)z^k/(q;q)_{2k+1}$; the pair $(G,H)$ realizes the Hahn--Exton
$q$-cosine/$q$-sine $J^{(3)}_{\mp1/2}$. The Casoratian $\Cas(G,\widetilde G)(z):=G(z)\widetilde
G(Qz)-G(Qz)\widetilde G(z)=c\,z^{1/2}$ is nonzero, and the exact $q$-Wronskian identity
\begin{equation}\label{eq:qwron}
   q\,G(z)\,H(Qz)-G(Qz)\,H(z)=q-1
\end{equation}
holds (verified to $40$ digits). We write $M$ for the associated rank-$2$ $\sigma$-module over $\C(z)$.

At $z=\infty$ the equation has a formal solution basis reflecting the two Newton slopes. With
$\theta_Q(x)=\sum_{n\in\Z}Q^{n(n-1)/2}x^n$, these are
\[
   f_1(z)=\theta_Q(-z)\,\hat g(z),\quad \hat g(z)=\sum_{k\ge0}b_kz^{-k},\ \ b_k=\frac{q^{\,k-k(k-1)/2}}{(q;q)_k};
   \qquad
   f_2(z)=\frac{h(z)}{\theta_Q(-z/q)},\quad h(z)=\sum_{k\ge0}c_kz^{-k},
\]
where $\hat g$ is $q$-Gevrey divergent (its $q$-Borel transform is $1/(q^2/\xi;q)_\infty$, so
$f_1$ is Borel--Laplace summable off one Stokes direction) and $h$ is convergent ($c_k=O(Q^{k^2/2})$).
The coefficients $b_k$, which drive the Galois argument below, satisfy
$b_k\asymp q^{-k^2/4}$; both $\hat g$ and $h$ are determined by \eqref{eq:Leq} and $b_0=c_0=1$. Full
details of the resummation are in \cite{Bonfioli2}; here only the asymptotic expansions are used.

\section{The difference Galois group}\label{sec:galois}

\begin{lemma}[irreducibility of the $\sigma$-module]
\label{lem:irreducible}
For $0<q<1$ the rank-$2$ $\sigma$-module of $G$ is irreducible over $\C(z)$: the Riccati equation
\[
   u(z)\,u(Qz)-a(z)\,u(z)+q=0,\qquad a=q+1-qz,
\]
has no solution $u\in\C(z)$.
\end{lemma}

\begin{proof}
Suppose $u=P/R$ with $P,R\in\C[z]$ coprime, of degrees $p,r$, leading coefficients $\alpha,\beta$.
Clearing denominators,
\[
   F(z)\;:=\;P(Qz)P(z)-a(z)P(z)R(Qz)+q\,R(z)R(Qz)\;\equiv\;0 .
\]
\emph{At $z=0$:} $F(0)=P(0)^2-(q+1)P(0)R(0)+qR(0)^2=0$; if $R(0)=0$ then $P(0)=0$, contradicting
coprimality, and symmetrically; so $P(0)R(0)\ne0$ and $u(0)\in\{1,q\}$.
\emph{Top degree:} the three terms have degrees $2p$, $p+r+1$, $2r$. If $p=r$, the top degree
$2p+1$ is attained by the middle term alone, with coefficient $q\alpha\beta Q^{r}\ne0$;
if $|p-r|\ge2$, the top degree is attained by the first or third term alone, with coefficient
$\alpha^2Q^{p}$ or $q\beta^2Q^{r}$. Both are impossible, so $|p-r|=1$.
\emph{Root pairing:} let $\zeta\ne0$ be a root of $R$ of multiplicity $m$, so $P(\zeta)\ne0$. At
$z=\zeta/Q$ the three terms of $F$ have orders $k$, $k+m+[\,a(\zeta/Q)=0\,]$, $m_0+m$, where
$k=\operatorname{ord}_{\zeta/Q}P$ and $m_0=\operatorname{ord}_{\zeta/Q}R$; since $F\equiv0$ the
minimum order cannot be attained by a single term, forcing $k\ge m_0+m\ge m$. Symmetrically, a root
$\xi\ne0$ of $P$ of multiplicity $k$ forces $\operatorname{ord}_{Q\xi}R\ge k$ (coprimality gives
$\operatorname{ord}_{Q\xi}P=0$, so the first term has order $k$ at $\xi$, the third
$\operatorname{ord}_{Q\xi}R$, and the middle at least $k+\operatorname{ord}_{Q\xi}R$; again no
single minimum). The two maps $\zeta\mapsto\zeta/Q$, $\xi\mapsto Q\xi$ are mutually inverse on the
nonzero root sets, and chains close at length two by coprimality. Summing multiplicities,
$p\le r\le p$, so $p=r$ --- contradicting $|p-r|=1$.
\end{proof}

\begin{theorem}[the difference Galois group contains $\mathrm{SL}_2$]
\label{thm:sl2}
For every $0<q<1$, the difference Galois group over $\C(z)$ of the module of $G$ contains
$\mathrm{SL}_2$.
\end{theorem}

\begin{proof}
For $N\ge1$ suppose $\operatorname{Sym}^N\!M$ has a rank-$1$ submodule, i.e.\ there is a nonzero
solution $w$ of the $N$-th symmetric-power equation with $r:=w(Q\,\cdot)/w\in\C(z)$. The products
$f_1^{N-k}f_2^{\,k}$ ($0\le k\le N$) form a solution basis (their Casoratian is a power
of $\operatorname{Cas}(f_1,f_2)\ne0$; checked numerically for $N=2$), so
$w=\sum_kc_kf_1^{N-k}f_2^k$ with constants $c_k$; set $k_0:=\min\{k:c_k\ne0\}$.

\emph{Case $k_0<N$.} Along a ray in a regular class, $f_2/f_1$ decays super-polynomially, so the
rational function $r$ has the Poincar\'e asymptotic expansion of the quotient of the dominant
component:
\[
   r\;\sim\;(-1/z)^{N-k_0}(-z/q)^{k_0}\,
   \Bigl(\tfrac{\hat g(Qz)}{\hat g(z)}\Bigr)^{N-k_0}
   \Bigl(\tfrac{h(Qz)}{h(z)}\Bigr)^{k_0}.
\]
The $j$-th coefficient of the formal series on the right is
$(N-k_0)\,b_jQ^{-j}\bigl(1+O(q^{cj})\bigr)$ with $b_j=q^{\,j-j(j-1)/2}/(q;q)_j$: the dominant term
comes from $b_jQ^{-j}$ in $\hat g(Qz)/\hat g(z)$, all competing products being smaller by a factor
$q^{cj}$ (verified: the ratio of the $j$-th coefficient to $(N-k_0)b_j(Q^{-j}-1)$ is
$1-10^{-6}$ at $j=12$, $q=\tfrac13$). These coefficients grow like $q^{-j^2/2}$, whereas a rational
function's expansion at $\infty$ has geometrically bounded coefficients. Since matching asymptotic
expansions are unique, no rational $r$ exists.

\emph{Case $k_0=N$.} Then $r=(-z/q)^N\rho^N$ with $\rho:=h(Q\,\cdot)/h$ meromorphic on $\C^*$;
$\rho^N\in\C(z)$ and single-valuedness force $\rho\in\C(z)$ (an $N$-th root of a rational function
is single-valued on $\C^*$ only if every divisor multiplicity, including at $0$ and $\infty$, is
divisible by $N$). Then $f_2(Qz)/f_2(z)=-(z/q)\rho(z)$ is rational, giving a rational solution of
the rank-$2$ Riccati equation --- contradicting Lemma~\ref{lem:irreducible}.

Hence no $\operatorname{Sym}^N\!M$ has a rank-$1$ submodule. $N=1$ is irreducibility; $N=2$
excludes the imprimitive (dihedral) case; and a projectively finite irreducible subgroup
($A_4,S_4,A_5$) fixes a line in some $\operatorname{Sym}^N$ with $N\le12$ (the classical binary
polyhedral invariants), which is likewise excluded. By the classification of algebraic subgroups of
$\mathrm{GL}_2$, the group contains $\mathrm{SL}_2$.
\end{proof}

\begin{proposition}[no $\sigma\delta$-integrability]
\label{prop:nonintegrable}
For every $0<q<1$ there is no $B\in\mathrm{gl}_2(\C(z))$ and no $c\in\C(z)$ with
$\theta(A)=\sigma(B)A-AB+cA$, where $A=\bigl(\begin{smallmatrix}0&1\\-q&a\end{smallmatrix}\bigr)$ and
$\theta=z\partial_z$.
\end{proposition}

\begin{proof}
Writing $B=\bigl(\begin{smallmatrix}\alpha&\beta\\\gamma&\delta\end{smallmatrix}\bigr)$, the entries give
$\gamma=-q\beta_1$, $\delta=\alpha_1+a\beta_1-c$, and eliminating $\alpha$ (the central terms $c$
cancel identically; verified symbolically) leaves the necessary condition
\[
   qa\,\beta_3\;=\;a_1(aa_1-q)\,\beta_2\;-\;a(aa_1-q)\,\beta_1\;+\;qa_1\,\beta\;+\;q\,z\bigl(a_1+q^2a\bigr),
\]
$\beta_k=\beta(Q^kz)$, $a=a(z)$, $a_1=a(Qz)$. A polynomial part of degree $m\ge0$ contributes top
coefficient $\propto\beta_m(1-q^{2+2m})\ne0$ at degree $m+3$, above the inhomogeneity: so $\beta$
has no polynomial part. If $\beta$ has a pole off the orbit $z_aQ^{\Z}$, the maximal-modulus pole
appears in exactly one term (the slot $z=\zeta Q^{-3}$, coefficient $qa\ne0$ there), and the
minimal-modulus pole likewise (slot $z=\zeta'$, coefficient $qa_1$): both impossible; so the poles
lie in the window $\{z_aQ^{-1},\dots,z_aQ^{3}\}$ ($a(z_aQ^k)=(q+1)(1-Q^k)$ vanishes only at $k=0$).
Sweeping the slots $z=z_aQ^{-4},\dots,z_aQ^{-1}$ kills the orders at $z_aQ^{-1},\dots,z_aQ^{2}$ one
by one (each pole appears in a single term with nonvanishing coefficient), and the order at
$z_aQ^{3}$ is killed at the slot $z=z_aQ$ --- coefficient $a_1(aa_1-q)$ with
$(aa_1-q)(z_aQ)=(q+1)^2(1-q^2)(1-q^4)-q$ --- or, where that vanishes (one root
$q\approx0.78027$), at the slot $z=z_aQ^{2}$ --- coefficient $a(aa_1-q)$ with
$(aa_1-q)(z_aQ^2)=(q+1)^2(1-q^4)(1-q^6)-q$ (one root $q\approx0.86753$); the two polynomials are
coprime, so for every $q\in(0,1)$ at least one slot applies. Hence $\beta\equiv0$, contradicting
$z(a_1+q^2a)=z\bigl[(q+1)(1+q^2)-2q^3z\bigr]\not\equiv0$. (Cross-check: at $q=\tfrac13$ the exact
linear system for poles of order $\le3$ in the window has $\operatorname{rank}(A)=15$,
$\operatorname{rank}[A|b]=16$.)
\end{proof}

\begin{corollary}[differential transcendence of $G$]
\label{cor:hypertranscendence}
For every $0<q<1$ the function $G(q,\cdot)$ satisfies no algebraic differential equation over
$\C(z)$: the Hahn--Exton $q$-cosine is differentially transcendental.
\end{corollary}

\begin{proof}[Proof, via parametrized difference-Galois theory]
We apply the parametrized ($\sigma,\theta$)-difference Galois theory with derivation
$\theta=z\partial_z$. The determinant equation $\sigma(y)=\det(A)\,y=q\,y$ has the solution
$y=z^{1/2}$ (indeed $\sigma(z^{1/2})=(Qz)^{1/2}=q\,z^{1/2}$, as $Q=q^2$), which is algebraic over $\C(z)$; hence the parametrized difference Galois group of $\sigma(y)=\det(A)y$
lies in $\C^\times$. This is exactly the hypothesis of Dreyfus--Hardouin--Roques
\cite[Prop.~2.9]{DHR2018}, which for a rank-$2$ system whose difference Galois group contains
$\mathrm{SL}_2$ (here Theorem~\ref{thm:sl2}) gives the dichotomy: the parametrized group
$\mathrm{Gal}^\theta$ either is conjugate into $\mathrm{GL}_2(\C)$ (all solutions differentially
algebraic) or contains $\mathrm{SL}_2(\widetilde{\mathcal C})$ (all nonzero solutions
differentially transcendental), and the \emph{first} alternative holds if and only if there exists
$B\in\mathrm{gl}_2(\C(z))$ with $\theta(A)=\sigma(B)A-AB$ (their integrability condition~(7);
$\theta(\det A)=0$ since $\det A=q$ is constant in $z$, so no trace term appears). By
Proposition~\ref{prop:nonintegrable} no such $B$ exists --- indeed none exists even allowing a
scalar twist $cA$. Therefore the second alternative holds:
$\mathrm{Gal}^\theta\supseteq\mathrm{SL}_2(\widetilde{\mathcal C})$, and every nonzero solution,
in particular $G(q,\cdot)$, is differentially transcendental over $\C(z)$.
\end{proof}


\section{The connection constant and the zero product}\label{sec:zeroproduct}

The behaviour of $G$ at $z=\infty$ is governed by a theta-comparison. Let
$\theta_Q(x)=\sum_{n\in\Z}Q^{n(n-1)/2}x^n$, so $\theta_Q(Qx)=x^{-1}\theta_Q(x)$ and, by the Jacobi
triple product, $\theta_Q(-z)=(Q;Q)_\infty(z;Q)_\infty(Q/z;Q)_\infty$, an entire function of order $0$
with simple zeros exactly at $z\in Q^{\Z}$. The following theorem evaluates the leading asymptotic
constant of $G$ and, as a consequence, the regularized product of its zeros; the constant
$C_1:=\lim_{R\to\infty}G(-R)/\theta_Q(R)$ coincides with the Stokes-invariant connection coefficient of
$G$ against the theta solution.

\begin{theorem}[the connection constant, and the zero product of the Hahn--Exton $q$-cosine]
\label{thm:zeroproduct}
For $0<q<1$,
\[
   C_1=\frac1{(q;q)_\infty},
   \qquad\text{and consequently}\qquad
   \prod_{k\ge1} z_k(q)\,Q^{\,k-1}\;=\;(q;q^2)_\infty ,
\]
where $z_1<z_2<\cdots$ are the positive zeros of $G(q,\cdot)$. Likewise, for the second solution's
underlying entire function $H$ with zeros $w_1<w_2<\cdots$,
$\prod_{k\ge1}w_k(q)\,q\,Q^{\,k-1}=(q^3;q^2)_\infty$. \textup{(Verified at three bases including
$q^\ast$, to $41$, $23$, $26$ digits for $C_1$ and $58$ digits for the product.)}
\end{theorem}

\begin{proof}
\emph{(1) An exact identity.} From $(q;q)_{2k}=(q;q)_\infty/(q^{2k+1};q)_\infty$ and Euler's
expansion $(x;q)_\infty=\sum_{j\ge0}(-1)^jq^{j(j-1)/2}x^j/(q;q)_j$ at $x=q^{2k+1}$, together with
$k(k-1)+2kj=(k+j)(k+j-1)-j(j-1)$ and the substitution $m=k+j$, absolute convergence of the double
sum permits the rearrangement
\[
   (q;q)_\infty\,G(z)\;=\;\sum_{m\ge0}(-1)^mq^{m(m-1)}z^m\,P_m(1/z),
   \qquad
   P_m(w):=\sum_{j=0}^m c_jw^j,\quad c_j=\frac{q^{\,j-j(j-1)/2}}{(q;q)_j}
\]
(the $c_j$ coincide with the $b_j$ of \S\ref{sec:prelim}; identity checked symbolically
through $z^8$ and numerically to $41$ digits).

\emph{(2) The constant.} Fix $u<0$ and put $z=uQ^{-N}$. The moduli $q^{m(m-1)}|z|^m$ peak at
$m_0=N+O(1)$ and decay like $q^{(m-m_0)^2+O(1)}$ away from it. For $1\le j\le m\le 2N$,
$2Nj-j(j-1)/2\ge Nj$, so $|c_jz^{-j}|\le Cq^{Nj}|u|^{-j}$ and
$|P_m(1/z)-1|\le C'q^{N}$ uniformly; the terms with $m>2N$ contribute a relative
$O(q^{N^2})$. The negative-$m$ part of $\theta_Q(-z)$ is $O(|z|^{-1})$, negligible against the peak.
Hence $(q;q)_\infty G(z)/\theta_Q(-z)\to1$ along $z=uQ^{-N}$, for each fixed $u<0$.

\emph{(3) Geometric zero dressing.} On the circle $|z-Q^{1-k}|=\varepsilon q^kQ^{1-k}$ the identity
of step (1) and the bounds of step (2) give
$|(q;q)_\infty G-\theta_Q(-z)|\le Cq^k|\theta_Q(-z)|_{\text{peak}}$, while $\theta_Q(-z)$ has a
simple zero at $Q^{1-k}$ with derivative bounded below there; Rouch\'e gives
$z_k=Q^{1-k}(1+O(q^k))$, i.e.\ $\log(z_kQ^{k-1})=O(q^k)$, absolutely summable.

\emph{(4) The product.} $G(z)=\prod_k(1-z/z_k)$ (order $0$, genus $0$, $G(0)=1$) and
$\theta_Q(-z)=(Q;Q)_\infty(z;Q)_\infty(Q/z;Q)_\infty$ (Jacobi). Along $z=-R$, $R\to\infty$,
\[
   \frac{G(-R)}{\theta_Q(R)}
   =\frac1{(Q;Q)_\infty(-Q/R;Q)_\infty}\prod_{k\ge1}\frac{1+R/z_k}{1+RQ^{k-1}} ,
\]
and for $x,y>0$ the function $R\mapsto\log\frac{1+Rx}{1+Ry}$ is monotone from $0$ to $\log(x/y)$, so
each factor is dominated by $|\log(z_kQ^{k-1})|$, summable by step (3). Dominated convergence gives
$G(-R)/\theta_Q(R)\to1/\bigl[(Q;Q)_\infty\prod_kz_kQ^{k-1}\bigr]$ as $R\to\infty$; evaluating along
the subsequence $R=|u|Q^{-N}$ by step (2), the limit equals $1/(q;q)_\infty$, whence
$C_1=1/(q;q)_\infty$ and $\prod_kz_kQ^{k-1}=(q;q)_\infty/(Q;Q)_\infty=(q;q^2)_\infty$ by Euler's
$(q;q)_\infty=(q;q^2)_\infty(q^2;q^2)_\infty$. The $H$-statement is identical with comparison
function $\theta_Q(-qz)$ and ratio $(1-q)/(q;q)_\infty$.
\end{proof}

\begin{remark}[the modular split at $z^\ast$]
\label{rem:zeroproduct-dioph}
At $q=q^\ast$ the product splits off the candidate rational zero:
\[
   \prod_{k\ge2}z_k(q^\ast)\,Q^{\,k-1}=\frac{(q;q^2)_\infty}{z^\ast}=0.988299884628746\ldots
   \quad(\text{verified to }30\text{ digits}).
\]
Were $z^\ast$ (equivalently $\beta_2$) algebraic, the modular value $(q;q^2)_\infty$ ---
transcendental at algebraic $q$ by Nesterenko's theorem \cite{Nesterenko} applied to the Euler
products --- would equal an algebraic number times the tail product $\prod_{k\ge2}z_kQ^{k-1}$. This is
consistent only if the tail product is itself transcendental; the arithmetic of the zero condition is
thereby confined to that tail, which is the content of the multiplicity problem of \cite{Bonfioli2}.
\end{remark}


\section{The arithmetic of the zeros}\label{sec:arith}

\subsection{Conjecture Z and its confluent evidence}
Conjecture~\ref{conj:Z} is the $q$-analogue of the classical theorem (Siegel; see also Lindemann for
the endpoint) that the positive zeros $j_{\nu,k}$ of the Bessel function $J_\nu$ are transcendental for
rational $\nu$. Two structural facts support it. First, Theorem~\ref{thm:main-transc}: $G$ is
differentially transcendental, so its zeros are not the zeros of any algebraic-differential shadow
whose arithmetic could be controlled classically. Second, in the confluent limit $q\to1$ the equation
\eqref{eq:Leq} degenerates to Bessel's equation and $z_k(q)\to(qj_{-1/2,k})^2$, so
Conjecture~\ref{conj:Z} deforms Siegel's theorem.

That confluence also explains, in one line, why the classical case is provable and the $q$-case is
not. Write $D_N$ for the denominator of the $N$-th partial sum. The classical engine (Lambert,
Lindemann, and every factorial-denominator proof) runs on a \emph{divergent step ratio}:
$D_{N+1}/D_N=N+1\to\infty$, so one further term is gained at asymptotically no arithmetic cost and
$\text{tail}\cdot D_N\asymp2/(N+1)\to0$ (measured: $0.310,\,0.163,\,0.099$ at $N=3,6,10$ for
$e=\sum1/k!$). The $q$-deformation replaces the factorial by $(q;q)_k$, whose step ratio is
\emph{constant} in the relevant scale: each further term gains $\asymp q^{2N}$ but costs
$\asymp t^{2N}$, and at $s=1$ the two cancel identically, so $\text{tail}\cdot D_N$ diverges
(measured at $q=\tfrac13$: $7.1,\,7.5\cdot10^5,\,5.1\cdot10^{14}$ at $N=2,4,6$). The bridge is
explicit: $(q;q)_k/\bigl((1-q)^kk!\bigr)\to1$ as $q\to1$ (measured $0.472,\,0.928,\,0.993$ at
$q=0.9,\,0.99,\,0.999$). The factorial --- and with it the divergent step ratio that powers
Lindemann --- exists only in the confluent limit; at any $q<1$ the ratio is constant, and a constant
ratio can never win. This is the same exactness of Remark~\ref{rem:thetabarrier} seen dynamically,
and it identifies precisely what a proof of Conjecture~\ref{conj:Z} must supply: an arithmetic
mechanism with no classical shadow.

\subsection{The self-referential specialization: $\beta_2$}
The reflection-group growth constant $\beta_2$ is the reciprocal square root of the least positive root
$q^\ast$ of $S(q)=G(q,2q(1-q))$, which is exactly the condition that the catalytic point
$z^\ast=2q^\ast(1-q^\ast)$ be a zero of $G(q^\ast,\cdot)$. Thus $\beta_2$ is governed by a
\emph{self-referential} specialization of Conjecture~\ref{conj:Z}: the zero and the base move together
along the diagonal $z=2q(1-q)$.

At this point the $q$-Wronskian \eqref{eq:qwron} yields a clean exact relation between two Hahn--Exton
values: since $G(q^\ast,z^\ast)=0$,
\begin{equation}\label{eq:wron-zero}
   G(q^\ast,\,q^{\ast2}z^\ast)\cdot H(q^\ast,\,z^\ast)=1-q^\ast
\end{equation}
(verified to $25$ digits) --- a product of two $q$-Bessel values at the algebraic point $z^\ast$ equal
to the algebraic number $1-q^\ast$. Together with $G(q^\ast,z^\ast)=0$ and the modular zero product of
Theorem~\ref{thm:zeroproduct-intro}, this places three exact constraints on the values at $z^\ast$; a
transcendence proof requires one more independent constraint than the reductio supplies, which is the
precise sense in which Conjecture~\ref{conj:Z} is a multiplicity (rather than a single-relation)
problem.

\subsection{The $q$-Siegel--Shidlovskii margin}
The natural route to Conjecture~\ref{conj:Z} is the $q$-analogue of the Siegel--Shidlovskii method
developed by Amou--Matala-aho--V\"a\"an\"anen \cite{AMV}. That method applies to Poincar\'e-type
$q$-difference systems --- a hypothesis \eqref{eq:Leq} satisfies at $z=0$ --- under a \emph{Siegel
margin} $\gamma<\Gamma$, where $\gamma$ is the ratio of the logarithmic height of the cleared
denominators to that of the analytic decay. For the Hahn--Exton class this ratio is
\[
   \gamma=\frac{\log(t/s)}{2\log t}\le\frac12\qquad(q=s/t\text{ in lowest terms}),
\]
the factor $2$ being the double Pochhammer $(q;q)_{2k}$; this is below the threshold, so the method does
not apply as stated. Extending it past the margin for this irregular-at-$\infty$ class is the
outstanding problem; the connection data needed for the irregular direction is available in closed form
\cite{Bonfioli2}.

\subsection{Integrality of the zero lattice}
The zeros, viewed as functions of the base, are integer power series after the Newton-polygon
rescale --- all of them, not only the first.

\begin{theorem}[integrality of the zero lattice]\label{thm:integrality}
For every $k\ge1$ the rescaled $k$-th zero $u_k(q):=q^{2(k-1)}z_k(q)$ lies in $1+q\,\Z[[q]]$.
In particular
\[
   z_1(q)=1-q-q^3+q^4-q^5+2q^6-2q^7+4q^8-6q^9+8q^{10}-14q^{11}+21q^{12}-\cdots\ \in\ \Z[[q]] .
\]
\end{theorem}

\begin{proof}
From $g_{k'}=(-1)^{k'}q^{k'(k'-1)}/(q;q)_{2k'}$ and the exponent identity
$k'(k'-1)-2(k-1)k'+k(k-1)=(k'-k)(k'-k+1)$,
\[
   F_k(q,u):=q^{k(k-1)}\,G\bigl(q,\,u\,q^{-2(k-1)}\bigr)
   =\sum_{k'\ge0}(-1)^{k'}q^{(k'-k)(k'-k+1)}\frac{u^{k'}}{(q;q)_{2k'}} .
\]
The exponent $(k'-k)(k'-k+1)$ is a product of consecutive integers, hence nonnegative, and vanishes
exactly at $k'\in\{k-1,k\}$: each zero scale is carried by its own edge of the strictly convex
Newton polygon of $G$. Since $1/(q;q)_{2k'}\in\Z[[q]]$ (partitions into parts $\le2k'$),
$F_k\in\Z[[q]][[u]]$ with $F_k(0,u)=(-1)^{k-1}u^{k-1}(1-u)$; at $u=1$ the value vanishes and
$\partial_uF_k(0,1)=(-1)^k$ is a unit, so Hensel's lemma in the complete local ring $\Z[[q]]$
yields a unique root $u_k\in1+q\,\Z[[q]]$, which by the scale localization
$z_k=q^{2(1-k)}(1+O(q^k))$ of \S\ref{sec:zeroproduct} is the $k$-th zero. (Verified to order
$q^{300}$ for $k\le4$: $u_2=1-q^6-2q^8+q^9-\cdots$, $u_3=1-q^{15}-\cdots$,
$u_4=1-q^{28}-\cdots$. The onset law visible here is proved next.)
\end{proof}

\subsection{Lattice values and the hexagonal law}
The deviation $u_k-1$ begins exactly at the $k$-th hexagonal number $k(2k-1)$. This reduces to, and
is proved by, an exact evaluation of $G$ at its own lattice points.

\begin{lemma}[parameter--argument swap]\label{lem:swap}
For $|Q|<1$,
$(b;Q)_\infty\,{}_1\phi_1(0;b;Q,z)$ is symmetric in $(b,z)$; equivalently
\begin{equation}\label{eq:dressing}
   (q;Q)_\infty\,G(q,z)=\sum_{n\ge0}\frac{(-1)^nq^{n^2}}{(Q;Q)_n}\,(zQ^n;Q)_\infty .
\end{equation}
\end{lemma}

\begin{proof}
In $(b;Q)_\infty\,{}_1\phi_1(0;b;Q,z)=\sum_n(-1)^nQ^{\binom n2}z^n\,(bQ^n;Q)_\infty/(Q;Q)_n$
expand $(bQ^n;Q)_\infty=\sum_i(-1)^iQ^{\binom i2}(bQ^n)^i/(Q;Q)_i$ (Euler); since
$\binom n2+\binom i2+ni=\binom{n+i}2$, the double sum is
$\sum_{n,i}(-1)^{n+i}Q^{\binom{n+i}2}z^nb^i/\bigl((Q;Q)_n(Q;Q)_i\bigr)$, manifestly symmetric.
For \eqref{eq:dressing} take $b=q$ and use $(z;Q)_\infty/(z;Q)_n=(zQ^n;Q)_\infty$ on the swapped
side, with $Q^{\binom n2}q^n=q^{n^2}$. (Verified to $40$ digits at generic $(b,z)$.)
\end{proof}

\begin{proposition}[square vanishing at the lattice]\label{prop:latticevalues}
For every $m\ge0$,
\[
   G\bigl(q,Q^{-m}\bigr)
   =(-1)^{m+1}\,\frac{(Q;Q)_\infty}{(q;Q)_\infty}
    \sum_{r\ge0}\frac{(-1)^r\,q^{(m+1+r)^2}}{(Q;Q)_r\,(Q;Q)_{m+1+r}} ,
\]
so $\operatorname{ord}_qG(q,Q^{-m})=(m+1)^2$ exactly, with leading coefficient $(-1)^{m+1}$.
\end{proposition}

\begin{proof}
Evaluate \eqref{eq:dressing} at $z=Q^{-m}$. The factor $(Q^{n-m};Q)_\infty$ vanishes for $n\le m$
and equals $(Q;Q)_\infty/(Q;Q)_{n-m-1}$ for $n\ge m+1$; reindexing $n=m+1+r$ gives the display.
The $q$-order of the $r$-th term is $(m+1+r)^2$, strictly increasing in $r$, and all other factors
are units in $\Z[[q]]$, so no cancellation can occur and the $r=0$ term decides. (Verified as an
identity in $\Z[[q]]$ to order $250$ for $m\le4$, and to $50$ digits at $q=0.3$.)
\end{proof}

\begin{corollary}[hexagonal law]\label{cor:hexagonal}
For every $k\ge1$,
\[
   u_k(q)=1-q^{\,k(2k-1)}\bigl(1+O(q)\bigr):
\]
the deviation of the $k$-th rescaled zero from $1$ begins exactly at the $k$-th hexagonal number,
with leading coefficient $-1$.
\end{corollary}

\begin{proof}
$F_k(q,1)=q^{k(k-1)}G(q,Q^{1-k})=(-1)^kq^{k(2k-1)}(1+O(q))$ by
Proposition~\ref{prop:latticevalues} at $m=k-1$ (note $k(k-1)+k^2=k(2k-1)$), while
$\partial_uF_k(q,1)=(-1)^k(1+O(q))$. The first Newton correction from $u=1$ is therefore
$-q^{k(2k-1)}(1+O(q))$, and all subsequent corrections are $O(q^{2k(2k-1)})$. This proves the
onsets $1,6,15,28$ observed above and predicts $u_5=1-q^{45}-\cdots$, confirmed directly.
\end{proof}

\begin{remark}\label{rem:latticevalues}
Three readings of Proposition~\ref{prop:latticevalues}. (a) The $r$-sum is again a Hahn--Exton
series, now of \emph{integer} order $m+1$ in base $Q$: the $q$-cosine ($\nu=-\tfrac12$) sampled
on its own zero lattice reproduces the integer-order family --- an order reflection
$-\tfrac12\mapsto m+1$. (b) Along any backward orbit $z=Q^{-m}w$, \eqref{eq:dressing} gives
$G(Q^{-m}w)\sim(Q^{-m}w;Q)_\infty/(q;Q)_\infty$ as $m\to\infty$, with constant
$(w;Q)_\infty(Q/w;Q)_\infty/(q;Q)_\infty=\theta_Q(-w)/(q;q)_\infty$ by the triple product ---
recovering the connection constant of the backward normal form independently. (c) At any zero
$z_k(q)$ of $G$, \eqref{eq:dressing} transfers the vanishing into the \emph{parameter} slot:
${}_1\phi_1\bigl(0;z_k(q);Q,q\bigr)=0$. At the crossing $z_1(q^\ast)=2q^\ast(1-q^\ast)$ of the
reductio this is a ${}_1\phi_1$-vanishing with all-rational parameters; its truncation ledger has
the same heights as the original (the swap preserves the exactness of
Remark~\ref{rem:thetabarrier}).
\end{remark}

\begin{remark}[arithmetic and analytic features of the zero curve]\label{rem:zerocurve-arith}
(a) The series is not rational and satisfies no algebraic relation $P(q,z_1)=0$ of bidegree
$\le(6,5)$ (exact linear algebra on $57$ coefficients). (b) Its radius of convergence is
$|q_c|=0.5545\ldots$, set by the real fold point of item (e); in particular
$q^\ast=0.4494\ldots$ lies inside the disc of convergence. (c) The
integrality opens a \emph{linear-height} approximation ledger: at $q=s/t$ the $N$-th partial sum is
rational with denominator $t^N$ and error $\asymp(q/R)^N$, $R=0.5545\ldots$, so the Liouville
criterion for the crossing $z_1(q)=2q(1-q)$ reads $s<R$ --- failing, but now only by the factor
$s/R\approx1.80$ at $s=1$, the narrowest margin among all ledgers recorded for this problem; the
obstruction has become the sub-unit radius, i.e.\ the complex collision of zeros. (d) The integer
coefficient sequence $1,-1,0,-1,1,-1,2,-2,4,-6,8,-14,21,-34,56,-88,148,-242,398,-669,1109,\dots$
appears to be new.

(e) The radius-limiting singularity is located and characterized: it is the simple fold
$q_c=-0.5545786101465797\ldots$, $z_c=2.5239525204417103\ldots$ (both to $230$ digits in the
repository), where the analytically continued branch collides with a second zero branch on the
negative $q$-axis ($G=G_z=0$ with $G_{zz}=0.2143\ldots\ne0$, $G_q=-1.5012\ldots\ne0$, so
$z_1(q)=z_c\pm\tilde C\sqrt{q-q_c}+\cdots$ with $\tilde C^2=2G_q/(-G_{zz})$ at the fold). The
fold law now has a closed-form constant:
\[
   |c_n|\,n^{3/2}|q_c|^n\ \longrightarrow\ \frac{|\tilde C|\sqrt{|q_c|}}{2\sqrt\pi}
   =0.7863083135\ldots,
\]
confirming $R=|q_c|$ exactly (the sequence reaches $0.78295$ at $n=299$ with a clean $1/n$ tail
extrapolating to $0.78621$; an earlier reading $0.769$ from $n\le120$ was preasymptotic). Neither
$q_c$ nor $z_c$ satisfies an integer polynomial relation of degree $\le8$ with coefficients up to
$10^{14}$, nor does the pair satisfy a bilinear one (PSLQ at $230$ digits, tolerance $10^{-200}$;
degree-$8$ candidates returned at $120$ digits were exposed as spurious by re-evaluation at $230$).

(f) No change of variable rescues the ledger of (c): if $q=\psi(u)$ with $\psi\in u\,\Z[[u]]$ (as
required to keep $z_1\circ\psi\in\Z[[u]]$) and $u^\ast=\varphi(q^\ast)=s'/t'\in\Q$, the criterion
again reads $|u^\ast|\,t'=s'<R_u$; but a non-polynomial integer power series has radius at most $1$,
so $R_u\le1$ while $s'\ge1$. Every integral uniformization is capped at the boundary case
$s'=1=R_u$: the sub-unit radius can be moved but not beaten. This is the geometric face of the same
boundary on which the theta and Rogers--Ramanujan series sit in the truncation ledger
(Remark~\ref{rem:thetabarrier}).

(g) The cap has an elementary final form, the \emph{discreteness cap}: for any non-polynomial
$X\in\Z[[q]]$, any $q_0=s/t$ inside its disc of convergence (away from the boundary), and any $N$,
the truncation tail satisfies $|X(q_0)-X_N(q_0)|=|c_M|(s/t)^M(1+o(1))\ge s^M\,t^{-M}(1-o(1))$, where
$M>N$ is the first index with $c_M\ne0$ --- at least the Liouville floor $t^{-M}$ of the truncation's
denominator whenever $s\ge1$, because a nonzero integer coefficient has $|c_M|\ge1$. (Measured on
$z_1$ at $q=\tfrac17$: tail/floor $=1.53,\,11.6,\,70.9,\,541$ at $N=6,10,14,18$.) Integrality thus
caps its own ledger: the property that grants linear heights also forbids the tail from undercutting
the floor. (h) The arithmetic class of $z_1$ is now pinned from above as well: it is \emph{not holonomic}. No
linear differential equation $\sum_{i\le r}P_i(q)z_1^{(i)}=0$ with $\deg P_i\le d$ exists for
$(r,d)$ up to $(8,10)$, $(6,14)$, $(5,12)$ (exact linear algebra on $110$--$116$ of the integer
coefficients, computed to order $120$; every coefficient matrix has full column rank). Thus $z_1$ is
an integer power series that is simultaneously non-algebraic, non-holonomic, and of radius $<1$ ---
precisely the class for which no arithmetic technology exists: the Chudnovsky--Andr\'e--Bost theory
and the recent arithmetic holonomy bounds of Calegari--Dimitrov--Tang, which are built for
``integrality plus structure'', all require holonomy, while P\'olya--Carlson requires radius exactly
$1$. Integrality alone, without holonomy, is not currently actionable.

(i) One exact relation ties the lattice across the bases $q$, $-q$, $q^2$: applying the zero
product of Theorem~\ref{thm:zeroproduct-intro} at the three bases and using
$(q;q^2)_\infty(-q;q^2)_\infty=(q^2;q^4)_\infty$,
\[
   \prod_{k\ge1}\varphi_k=1,\qquad
   \varphi_k:=\frac{u_k(q)\,u_k(-q)}{u_k(q^2)} .
\]
The relation is not termwise: $\varphi_1=1+2q^6+4q^8+8q^{10}+18q^{12}+\cdots$,
$\varphi_2=1-2q^6-\cdots$, $\varphi_3=1+2q^{20}+\cdots$, $\varphi_4=1-2q^{28}-\cdots$ (order
$300$; adjacent leading defects cancel in pairs). No Mahler structure underlies it. With $2000$
coefficients computed modulo a prime (a primitive integer relation survives reduction, so full rank
modulo one prime excludes a rational relation outright), $u_1$ satisfies no polynomial relation
$P\bigl(q,u_1(q),u_1(q^d)\bigr)=0$ for $d=2,3$ at bidegrees up to $(490,1,1)$, $(150,2,2)$,
$(100,3,3)$, $(60,4,4)$, $(40,5,5)$, $(30,6,6)$, $(24,7,7)$, $(20,8,8)$; no order-two system
relation $P\bigl(q,u_1(q),u_1(q^2),u_1(q^4)\bigr)=0$ at $(30,2,2,2)$, $(20,3,3,3)$; no
difference--differential mix $P\bigl(q,u_1,\theta u_1,u_1(q^2)\bigr)=0$ at the same profiles; no
quasi-modular coupling: $\{1,\;\theta\!\log u_1(q),\;\theta\!\log
u_1(q^2),\;E_2(q),\;E_2(q^2),\;E_4(q)\}$ admits no $\Q(q)$-linear relation with polynomial
coefficients of degree up to $399$ (so $u_1$ is not a hidden eta-quotient or $E_2$-ring object of
level $2$); no linear relation among $\{1,\log u_1(q),\log u_1(q^2)\}$ to degree $665$; and no
Mahler relation for the logarithmic derivative $\theta\!\log u_1$ itself up to $(150,2,2)$,
$(100,3,3)$, $(60,4,4)$. Earlier and independently: the system $\{u_1,u_2,u_3\}$ admits no
bilinear-profile relation ($300$ exact coefficients), and no zero of the base-$q^2$ cosine or sine
is a rational multiple of $q^{e}z_i(q)^{a}w_j(q)^{b}$ for $|e|\le10$, $a+b\le2$ ($60$ digits, two
bases). Mahler's method --- intrinsically $q$-analytic, with no classical shadow --- thus finds no
foothold on the zero curve at any accessible complexity.

(j) The fold is the natural point for a rationality reductio, because it is the unique point of
the $(q,z)$-plane carrying \emph{two} vanishings ($G=\theta G=0$; a triple point $G=\theta
G=\theta^2G=0$ is codimension three in two variables, hence absent). The first-jet web there
consists of six values, $A=G(Qz_c)$, $B=\theta G(Qz_c)$, $h=H(z_c)$, $h_1=H(Qz_c)$, $\eta=\theta
H(z_c)$, $\eta_1=\theta H(Qz_c)$, where $H$ is the sine-side companion, which satisfies the mirror
equation $H(z)-a(z)H(Qz)+qH(Q^2z)=0$ (the $z^{1/2}$ in the second solution $z^{1/2}H$ moves the
outer $q$). Exactly two finite-height relations hold: the Casoratian at $z_c$, $Ah=1-q_c$, and its
$\theta$-derivative, $Bh=-A\eta$ (both verified to $236$ digits). For any mixed jet-Casoratian
$C_{u,v}(z)=qu(z)v(Qz)-u(Qz)v(z)$ with $L(u)=f$, $\widetilde L(v)=\tilde g$ the $a$-terms cancel
in the increment:
\[
   C_{u,v}(Qz)-C_{u,v}(z)=\tilde g(z)\,u(Qz)-f(z)\,v(Qz),
\]
so every telescoped sum of first-jet orbit bilinears closes back onto a Casoratian. In particular
$B\eta=\sum_{m\ge0}qQ^mz_c\bigl[G\,\theta H-H\,\theta G\bigr](Q^{m+1}z_c)$ (verified to
$121$ digits) is exact but carries an infinite rational series as coefficient --- height-infinite.
The ledger of a point evaluation is therefore fixed: each further $\theta$- or
$\delta$-jet level adds at least three new values and exactly one finite-height relation, so the
deficit (four at the fold, five at a generic zero) cannot be closed by jet-raising, and the fold's
two vanishings are the maximum the plane admits. Point evaluation alone, at any point including the
fold, cannot complete a rationality reductio for this module.

Together, (c), (f), (g) and Remark~\ref{rem:thetabarrier} express one law from four sides:
for every representation of this problem yet constructed, the arithmetic cost equals the analytic
decay \emph{exactly}, with zero excess; a proof of Conjecture~\ref{conj:Z} must produce strict
excess, and rearrangement, reparametrization, or symmetrization of the object's own
$\Q(q)$-structure provably cannot.
\end{remark}

\subsection{The second-kind forms and the exact reduction}
The $q$-Siegel construction for this class is explicit. For $N\ge1$ put
\[
   P_N(q,z):=(q;q)_{2N}\sum_{k=0}^{N}\frac{(-1)^kq^{k(k-1)}z^k}{(q;q)_{2k}}
   =\sum_{k=0}^{N}(-1)^kq^{k(k-1)}z^k\!\!\prod_{i=2k+1}^{2N}\!\!(1-q^i)\ \in\ \Z[q,z].
\]
\begin{proposition}[the second-kind forms]\label{prop:secondkind}
$\deg_qP_N=2N^2+N$ and $\deg_zP_N=N$ \textup{(exact; verified $N\le4$)}, and the remainder
$R_N(q,z):=(q;q)_{2N}G(q,z)-P_N(q,z)$ satisfies, uniformly for $q$ in compacts of $(0,1)$ and $z$
bounded,
\[
   |R_N(q,z)|=q^{\,N^2(1+o(1))} .
\]
\end{proposition}
\begin{proof}
Membership in $\Z[q,z]$ and $\deg_zP_N=N$ are clear from the displayed product form, since
$(q;q)_{2N}/(q;q)_{2k}=\prod_{i=2k+1}^{2N}(1-q^i)\in\Z[q]$. For the $q$-degree, the $k$-th summand has
degree $k(k-1)+\sum_{i=2k+1}^{2N}i=2N^2+N-k^2-2k$, maximal at $k=0$ with value
$\deg(q;q)_{2N}=2N^2+N$; the maximum is attained once, so no cancellation occurs. For the remainder,
$R_N=(q;q)_{2N}\sum_{k>N}(-1)^kq^{k(k-1)}z^k/(q;q)_{2k}$, whose leading term ($k=N+1$) is
$\pm q^{N(N+1)}z^{N+1}(q;q)_{2N}/(q;q)_{2N+2}$; the tail is dominated by a geometric series in
$q^{2N}$, so $|R_N|=q^{N^2+O(N)}$ uniformly. \textup{(Measured $\log|R_N|/N^2\to\log q$ at both
$z=2q(1-q)$ and a generic $z_0$.)}
\end{proof}

\begin{proposition}[reduction to the Siegel margin]\label{prop:reduction}
Let $q=s/t\in(0,1)$ and let $z_0=a/b\in\Q^\times$ with $G(q,z_0)=0$. Then $P_N(q,z_0)=-R_N(q,z_0)$ and
$t^{2N^2+N}b^N\,P_N(q,z_0)\in\Z$. Consequently, if for infinitely many $N$ one has $P_N(q,z_0)\ne0$ and
\[
   |R_N(q,z_0)|<t^{-(2N^2+N)}b^{-N},
\]
then $G(q,z_0)\ne0$ --- a contradiction. By Proposition~\ref{prop:secondkind} the displayed criterion
reads $q^{N^2}<t^{-2N^2}$ up to $O(N)$ in the exponent, i.e.
\[
   \log(t/s)>2\log t\iff st<1,
\]
which is false. The deficit is exactly the factor $2$ in $\deg_qP_N=\deg(q;q)_{2N}$: the \emph{double}
Pochhammer.
\end{proposition}
\begin{proof}
$P_N(q,z_0)=(q;q)_{2N}G(q,z_0)-R_N(q,z_0)=-R_N(q,z_0)$ by hypothesis; clearing denominators of
$P_N\in\Z[q,z]$ at $q=s/t$, $z=a/b$ gives $t^{\deg_qP_N}b^{\deg_zP_N}P_N(q,z_0)\in\Z$, and a nonzero
integer has modulus $\ge1$. The stated evaluation of the criterion is
Proposition~\ref{prop:secondkind}.
\end{proof}

\begin{corollary}[what would suffice]\label{cor:suffices}
Conjecture~\ref{conj:Z} at $z_0$ --- hence, on the diagonal, the irrationality of $\beta_2$ --- follows
from any family of forms $\Lambda_N\in\Z[q,z]$ with $\Lambda_N(q,z_0)\ne0$, remainder
$q^{N^2(1+o(1))}$ against $G$, and
\[
   \deg_q\Lambda_N\;\le\;N^2\,\frac{\log(t/s)}{\log t}\,(1-\varepsilon),
   \qquad\text{in particular }\deg_q\Lambda_N\le N^2(1-\varepsilon)\text{ when }s=1 .
\]
The forms of Proposition~\ref{prop:secondkind} have $\deg_q=2N^2$: it suffices to \emph{halve} the
$q$-degree at the same remainder, or equivalently to double the remainder decay to $q^{2N^2}$. A
single-Pochhammer clearing $(q;q)_N$ would have degree $\tfrac12N^2(1+o(1))$ and would suffice at once;
no such representation of $G$ along a non-$q$-power argument exists in the
Rogers--Ramanujan/Bailey framework.
\end{corollary}

\begin{remark}[the theta-decay barrier: why Corollary~\ref{cor:suffices} is unreachable by truncation]
\label{rem:thetabarrier}
The hypothesis of Corollary~\ref{cor:suffices} cannot be met by \emph{any} truncate-and-clear form.
Let $\sum_kc_k(q)q^{\alpha k^2}z^k$ ($\alpha>0$, $c_k\in\Q(q)$) be truncated at $N$ and cleared. The
$k=N$ summand already contributes $q^{\alpha N^2}$, so $\deg_q\ge\alpha N^2$; the remainder is the
first omitted term, of size $q^{\alpha N^2(1+o(1))}$. The criterion $|R_N|<t^{-\deg_q}$ therefore reads
\[
   \alpha N^2\log(t/s)\;>\;\alpha N^2\log t\iff s<1,
\]
false for every $q=s/t\in(0,1)$: \emph{the decay and the degree are driven by the same factor
$q^{\alpha k^2}$ and cancel exactly}. Denominators only make matters worse. Numerically, the cleared
truncations have $\deg_q=N^2$ exactly for the theta series $\sum q^{k^2}z^k$ and for the
Rogers--Ramanujan series $\sum q^{k^2}/(q;q)_k$ (both $\beta:=\deg_q/N^2=1$, the boundary), while the
Hahn--Exton series sits at $\beta\to2$ ($2.25,2.17,2.12$ at $N=4,6,8$) --- a further factor $2$ past a
boundary that is itself already unreachable.

This explains the shape of the known landscape rather than merely our position in it: the
transcendence of theta values and of Rogers--Ramanujan values --- both at $\beta=1$ --- is known
\emph{only} through Nesterenko's modular method \cite{Nesterenko}, never through a Siegel/Liouville
margin, precisely because no theta-decay series can cross $\beta<1$. Consequently the only mechanisms
available for Conjecture~\ref{conj:Z} are modularity, which $G$ provably lacks (it is off the modular
class; see \cite{Bonfioli2}), or a multiplicity theory for the associated $(\sigma,\delta)$-module. The
double Pochhammer is not the whole obstruction --- it is a factor $2$ on top of a structural one.
\end{remark}

\subsection{An unconditional effective bound}
Although Conjecture~\ref{conj:Z} is open, the specific root $q^\ast$ admits an effective
non-rationality statement, proved by certified interval arithmetic.

\begin{proposition}[effective non-rationality of $q^\ast$]
\label{prop:effective}
The least positive root $q^\ast$ of $S$ admits no rational representation $s/t$ with
$t\le10^{191}$. Consequently, if $\beta_2=u/v$ in lowest terms then $u>10^{95}$.
\end{proposition}

\begin{proof}
All steps are certified interval arithmetic (outward rounding; \texttt{mpmath.iv}; script
\texttt{beta2\_effective\_irrationality.py}), with the series tail bounded by
$(q;q)_{2k}\ge(1-2q)/(1-q)>0$ for $q<\tfrac12$ and term ratios $<\tfrac12$ beyond the truncation.
\textup{(i)} $S>0$ on $[0,0.40]$ ($400$ certified chunks). \textup{(ii)} $S'\le-1$ on $[0.40,0.46]$
($600$ certified chunks; $S'$ evaluated term-by-term with the logarithmic-derivative weights), so
$S$ has at most one zero in $[0.40,0.46]$. \textup{(iii)} with $a=0.d_1\cdots d_{385}$ the
$385$-digit truncation of $q^\ast$ and $b=a+2\cdot10^{-385}$, certified evaluation gives $S(a)>0>S(b)$;
by \textup{(i)}--\textup{(ii)}, $q^\ast\in[a,b]$. \textup{(iv)} the interval $[a,b]$ contains the
rational $s_0/t_0$ (Stern--Brocot minimal fraction) with $t_0\ge10^{192}$, verified by exact rational
comparison, and certified $S(s_0/t_0)\ne0$. If $m/n\in[a,b]$ with $n\le10^{191}$ and $m/n\ne s_0/t_0$,
then $|m/n-s_0/t_0|\ge1/(nt_0)\ge10^{-384}>b-a$, impossible; so $m/n=s_0/t_0$, whence
$n=t_0>10^{191}$, a contradiction. Thus no rational with denominator $\le10^{191}$ lies in $[a,b]$
at all, in particular none equals $q^\ast$. For $\beta_2=u/v$ in lowest terms,
$q^\ast=v^2/u^2$ in lowest terms, so $u^2>10^{191}$.
\end{proof}


\begin{thebibliography}{9}
\bibitem{AMV} M.~Amou, T.~Matala-aho, K.~V\"a\"an\"anen, \emph{On Siegel--Shidlovskii's theory for
$q$-difference equations}, Acta Arith.\ \textbf{127} (2007), 309--335.
\bibitem{ADR} C.~E.~Arreche, T.~Dreyfus, J.~Roques, \emph{Differential transcendence criteria for
second-order linear difference equations and elliptic hypergeometric functions}, J.~\'Ec.\ polytech.\
Math.\ \textbf{8} (2021), 147--168.
\bibitem{Bonfioli2} V.~Bonfioli, \emph{Transcendence of a planar reflection--group growth series and
its relaxed companion}, companion manuscript (2026); code and data at the repository above.
\bibitem{DHR2018} T.~Dreyfus, C.~Hardouin, J.~Roques, \emph{Hypertranscendence of solutions of Mahler
equations}, J.~Eur.\ Math.\ Soc.\ \textbf{20} (2018), 2209--2238.
\bibitem{HS} C.~Hardouin, M.~F.~Singer, \emph{Differential Galois theory of linear difference
equations}, Math.\ Ann.\ \textbf{342} (2008), 333--377.
\bibitem{KS} H.~T.~Koelink, R.~F.~Swarttouw, \emph{On the zeros of the Hahn--Exton $q$-Bessel function
and associated $q$-Lommel polynomials}, J.\ Math.\ Anal.\ Appl.\ \textbf{186} (1994), 690--710.
\bibitem{Nesterenko} Yu.~V.~Nesterenko, \emph{Modular functions and transcendence questions},
Sb.\ Math.\ \textbf{187} (1996), 1319--1348.
\end{thebibliography}

\end{document}
