\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[T1]{fontenc}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{question}[theorem]{Question}

\title{Universality and the right-angled Coxeter envelope\\ of orthoscheme reflection groups}
\author{Vico Bonfioli}
\date{}

\begin{document}
\maketitle

\begin{abstract}
\noindent The right-triangle orbit-growth universality of the companion papers is the planar case of a
general phenomenon. We prove an abstract dichotomy: for any family of quotients of a common group,
spherical growth is shape-independent through a horizon equal to half the shortest added relator and
first deviates exactly there. For orthoscheme (path-simplex) reflection groups the common group is the
\emph{right-angled Coxeter group} (RACG) on the orthoscheme's orthogonality graph, whose growth we
compute in closed form: it is rational, with rate the algebraic number $r_n=1+2\cos\frac{2\pi}{n+3}$ (the
spectral radius attached to the path $P_{n+1}$) and numerator $(1+t)^{\lfloor n/2\rfloor+1}$. The actual
geometric group $W_T\le\mathrm{Isom}(\R^n)$ is \emph{amenable}, hence a \emph{proper} quotient of the
non-amenable RACG, so its growth matches this rational envelope only up to a finite, shape-dependent
depth and then deviates strictly below. In the plane the deviation occurs at depth $10$, where the growth
becomes the transcendental sequence \textup{A396406}; in every higher dimension it lies beyond direct
computation. We are therefore \emph{not} entitled to conclude that higher dimensions are rational: the
finite envelope agreement observed for $n\ge3$ is the same pattern that, at $n=2$, precedes the
transcendental collapse. Whether the plane is exceptional, or every dimension is eventually
transcendental, is left as an explicit open problem.
\end{abstract}

\section{The abstract universality--deviation dichotomy}

Let $G=\langle S\rangle$ be a group with finite symmetric generating set $S$ and word length $\ell_G$.
For $N\trianglelefteq G$ write $Q=G/N$, $s_d(H)=\#\{h:\ell_H(h)=d\}$, and $K(N)=\min\{\ell_G(g):g\in
N\setminus\{1\}\}$.

\begin{theorem}[Universality--deviation dichotomy]
\label{thm:dichotomy}
With $n:=\lceil K(N)/2\rceil$: $\ s_d(Q)=s_d(G)$ for $d<n$, and $s_n(Q)<s_n(G)$.
\end{theorem}

\begin{proof}
$\pi:G\to Q$ is surjective and $1$-Lipschitz, so $s_d(Q)\le s_d(G)$. If $d<n$ and $\pi g=\pi h$
($g,h\in B_d(G)$) then $gh^{-1}\in N$ has length $\le2d<K(N)$, forcing $g=h$; the same bound excludes a
shorter representative, so $\pi$ is length-preserving on $B_d(G)$ and $s_d(Q)=s_d(G)$ for $d<n$. Splitting
a geodesic word for $g_0\in N$ with $\ell_G(g_0)=K(N)$ as $g_0=w_1w_2$, $\ell_G(w_1)=n$, gives two
distinct elements of $B_n(G)$ sharing a fiber, so $s_n(Q)<s_n(G)$.
\end{proof}

Applied to a family $\{Q_T=G/N_T\}$, this is universality: all members share $G$'s growth through their
horizon $n_T=\lceil K(N_T)/2\rceil$, then deviate. \textbf{The point of this paper is that for
orthoscheme reflection groups the ambient $G$ is a rational-growth RACG, but the members $Q_T$ are proper
quotients, so the shared rational growth is only a finite-depth envelope.}

\section{The orthoscheme family and its RACG envelope}

An $n$-orthoscheme has vertices $P_0,\dots,P_n$ with mutually orthogonal consecutive edges $P_{k-1}P_k=
\ell_k e_k$; its $n+1$ facet reflections $R_0,\dots,R_n$ generate $W_T\le\mathrm{Isom}(\R^n)$ (the
right-triangle group when $n=2$). Two facts hold for \emph{every} orthoscheme: $R_i^2=1$, and
$R_iR_j=R_jR_i$ when $|i-j|\ge2$ (orthogonal facets). For generic legs consecutive facets meet at an
irrational multiple of $\pi$, imposing no finite bond. The largest group compatible with these relations
is the \emph{right-angled Coxeter group}
\[
   W_n \;=\; \bigl\langle R_0,\dots,R_n \ \big|\ R_i^2=1,\ (R_iR_j)^2=1\ (|i-j|\ge2)\bigr\rangle,
\]
and there is a surjection $\rho\colon W_n\twoheadrightarrow W_T$.

Before computing the envelope we record that the ``generic growth'' is a well-defined object: it does
not depend on which generic shape one picks. Let $F=\ast_{i=0}^{n}\langle r_i\rangle$ be the free
product of $n+1$ copies of $\Z/2$, let $U=\{\ell=(\ell_1,\dots,\ell_n):\ell_i>0\}$ be the leg space
($\ell_1=1$), and let $\rho_\ell\colon F\to\mathrm{Isom}(\R^n)$, $r_i\mapsto R_i(\ell)$, be the
reflection representation; put $N_{\mathrm{gen}}=\bigcap_{\ell\in U}\ker\rho_\ell$.

\begin{theorem}[Generic rigidity]
\label{thm:rigidity}
For every $n\ge2$ there is a co-null subset $U_{\mathrm{gen}}\subseteq U$ --- the complement of a
countable union of proper real-algebraic subvarieties --- such that $\ker\rho_\ell=N_{\mathrm{gen}}$,
hence $W_\ell\cong F/N_{\mathrm{gen}}$, for every $\ell\in U_{\mathrm{gen}}$. Consequently the geodesic
growth $s_d(W_\ell)$ is constant on $U_{\mathrm{gen}}$, equal to $s_d(F/N_{\mathrm{gen}})$, and this
generic value is the maximum of $s_d(W_\ell)$ over all $\ell\in U$. Moreover $U_{\mathrm{gen}}$ contains
every $\ell$ whose coordinates are algebraically independent over $\Q$; in particular it is dense.
\end{theorem}

\begin{proof}
Each facet normal $m_j(\ell)$ is a fixed polynomial vector in $\ell$ (a cofactor of the vertex
differences, which are linear in $\ell$), so the reflection $R_j(\ell)=I-2\,m_jm_j^{\!\top}/(m_j\!\cdot\!m_j)$
(with its translation) has entries in $\Q(\ell)$ with denominator the nonvanishing polynomial
$m_j\!\cdot\!m_j$. Hence every entry of $\rho_\ell(v)=\prod_k R_{i_k}(\ell)$ lies in $\Q(\ell)$, and for
each $v\in F$ the locus $V_v=\{\ell\in U:\rho_\ell(v)=I\}$ --- the simultaneous vanishing of finitely
many rational functions, cleared of their nonvanishing denominators --- is Zariski-closed in the
irreducible variety $U$. If $v\notin N_{\mathrm{gen}}$ then $\rho_\ell(v)\ne I$ for some $\ell$, so $V_v$
is a \emph{proper} closed subvariety, of measure zero. As $F$ is countable,
$B:=\bigcup_{v\in F\setminus N_{\mathrm{gen}}}V_v$ is a countable union of proper subvarieties;
$U_{\mathrm{gen}}:=U\setminus B$ is co-null and dense. For $\ell\in U_{\mathrm{gen}}$ one has
$N_{\mathrm{gen}}\subseteq\ker\rho_\ell$ (universal relations hold everywhere) and
$\ker\rho_\ell\subseteq N_{\mathrm{gen}}$ (any $v\in\ker\rho_\ell\setminus N_{\mathrm{gen}}$ would place
$\ell\in V_v\subseteq B$), so $\ker\rho_\ell=N_{\mathrm{gen}}$ and $W_\ell\cong F/N_{\mathrm{gen}}$.
Isomorphic groups with the marked generators have identical geodesic growth, giving the constant value;
for any $\ell$, $\ker\rho_\ell\supseteq N_{\mathrm{gen}}$ makes $W_\ell$ a quotient of
$F/N_{\mathrm{gen}}$, so $s_d(W_\ell)\le s_d(F/N_{\mathrm{gen}})$, i.e.\ the generic value is the maximum.
Finally each $V_v$ is defined over $\Q$; if the coordinates of $\ell$ are algebraically independent over
$\Q$ then $\ell$ lies on no proper $\Q$-subvariety, so $\ell\notin B$, i.e.\ $\ell\in U_{\mathrm{gen}}$.
Such $\ell$ exist since $\dim U=n-1\ge1$.
\end{proof}

The non-consecutive commuting relations $(R_iR_j)^2=1$ ($|i-j|\ge2$) hold identically in $\ell$ (the
interior normals $m_j$ are supported on the two coordinates $\{j{-}1,j\}$, so non-adjacent normals are
orthogonal for \emph{all} legs; verified symbolically for $n\le5$), so they lie in $N_{\mathrm{gen}}$ and
$F/N_{\mathrm{gen}}$ is a quotient of the RACG $W_n$ below.

\begin{theorem}[The RACG envelope is rational]
\label{thm:envelope}
The spherical growth series of $W_n$ is rational,
\[
   W_n(t)\;=\;\frac{(1+t)^{\lfloor n/2\rfloor+1}}{N_n(t)},
   \qquad N_n(t)=(1+t)^{\,n+1}\,c_{\Gamma_n}\!\Bigl(\tfrac{-1}{1+t}\Bigr),
\]
where $c_{\Gamma_n}$ is the clique polynomial of the orthogonality graph $\Gamma_n$ (vertices $0,\dots,n$;
edges $|i-j|\ge2$). Its growth rate is $r_n=1+2\cos\frac{2\pi}{n+3}$, the largest root of the reciprocal
of $N_n$, and equals $1$ plus the spectral radius $2\cos\frac{2\pi}{n+3}$ of the path $P_{n+1}$. Explicitly
\[
\renewcommand{\arraystretch}{1.4}
\begin{array}{c|c|c}
n & W_n(t) & r_n\\\hline
2 & (1+t)^2/(1-t-t^2) & \varphi=1.618\ldots\\
3 & (1+t)^2/(1-2t) & 2\\
4 & (1+t)^3/(1-2t-t^2+t^3) & 1+2\cos\tfrac{2\pi}7=2.24698\ldots\\
5 & (1+t)^3/(1-3t+t^2+t^3) & 1+\sqrt2\\
6 & (1+t)^4/(1-3t+3t^3) & 1+2\cos\tfrac{2\pi}9=2.53209\ldots\\
7 & (1+t)^4/(1-2t-3t^2+4t^3-t^4) & \varphi^2=2.61803\ldots
\end{array}
\]
\end{theorem}

\begin{proof}
$W_n$ is a RACG, so its finite parabolics are exactly the cliques of $\Gamma_n$, each a group
$(\Z/2)^{|C|}$ with polynomial growth $(1+t)^{|C|}$. Steinberg's formula gives $1/W_n(t^{-1})=\sum_{C\
\mathrm{clique}}(-1)^{|C|}/(1+t)^{|C|}=c_{\Gamma_n}(-1/(1+t))$; inverting and substituting $t\mapsto t^{-1}$
yields the displayed rational form. $\Gamma_n$ is the complement of the path $P_{n+1}$, and the reciprocal
of $N_n$ is (a Chebyshev multiple of) the characteristic polynomial of $P_{n+1}$, whose largest eigenvalue
is $2\cos\frac{\pi}{n+2}$; a short computation with the independence polynomial of the path pins the
dominant growth root at $r_n=1+2\cos\frac{2\pi}{n+3}$. The tabulated entries are the cases $n=2,\dots,7$,
verified symbolically.
\end{proof}

\section{The geometric group is a proper amenable quotient}

\begin{proposition}[Finite-horizon envelope]
\label{prop:quotient}
For every $n\ge2$, $W_T$ is a \emph{proper} quotient of $W_n$; consequently, by
Theorem~\ref{thm:dichotomy}, the geometric growth $s_d(W_T)$ equals the rational envelope $[t^d]W_n(t)$ only
for $d<n_T:=\lceil K(\ker\rho)/2\rceil$, and $s_{n_T}(W_T)<[t^{n_T}]W_n(t)$.
\end{proposition}

\begin{proof}
$W_T\le\mathrm{Isom}(\R^n)=\R^n\rtimes O(n)$ is amenable (an extension of the amenable $\R^n$ by the compact
$O(n)$). The RACG $W_n$ is non-amenable: its orthogonality graph $\Gamma_n$ is not complete multipartite
(its complement, the path $P_{n+1}$, is connected and not a disjoint union of cliques for $n\ge2$), so
$W_n$ contains a free subgroup of rank $2$. An amenable group cannot be isomorphic to a non-amenable one,
so $\ker\rho\ne\{1\}$ and $W_T$ is a proper quotient. Theorem~\ref{thm:dichotomy} then gives the horizon.
\end{proof}

\begin{corollary}[The plane]
\label{cor:plane}
For $n=2$ the horizon is $n_{T}=10$: $s_d(W_T)=F_{d+3}=[t^d]W_2(t)$ for $d\le9$, and at $d=10$ the growth
deviates, becoming the sequence \textup{A396406}, which is transcendental over $\Q$ (companion paper: the
relaxed companion unconditionally; the true series conditionally on one amplitude estimate) with rate
$\beta_2=1.4916\ldots<\varphi=r_2$. This is the only dimension in which the deviation has been computed
past its onset.
\end{corollary}

\section{What is proved, and the open problem}

Four things are proved unconditionally and dimension-freely: the universality--deviation dichotomy
(Theorem~\ref{thm:dichotomy}); \emph{generic rigidity} (Theorem~\ref{thm:rigidity}), which makes ``the
generic $n$-orthoscheme growth'' a well-defined object $s_d(F/N_{\mathrm{gen}})$ realized by a co-null set
of shapes; the rationality of the RACG envelope with rate $r_n$ (Theorem~\ref{thm:envelope}); and,
structurally, that the geometric group is always a proper amenable quotient, so its growth is the envelope
only up to a finite horizon (Proposition~\ref{prop:quotient}).
For $3\le n\le7$ the geometric growth was computed to depth $13$--$16$ and agrees with the envelope
throughout that range; but Proposition~\ref{prop:quotient} guarantees it must eventually deviate below,
and the deviation depth $n_T$ for $n\ge3$ exceeds what we can reach.

The tempting conclusion --- that the geometric growth is rational for $n\ge3$, so the plane is the unique
transcendental dimension --- is \emph{not} justified. The evidence for it is precisely the finite envelope
agreement, and at $n=2$ that same agreement (through depth $9$) precedes the transcendental collapse. We
therefore state it as an open problem rather than a theorem.

\begin{question}[Eventual transcendence]
\label{q:transc}
For $n\ge3$, is the geometric growth series of the generic $n$-orthoscheme group rational, algebraic, or
transcendental? Equivalently: what is the arithmetic of the amenable quotient $W_n/\ker\rho$ past its
deviation horizon $n_T$? The plane is exceptional iff the answer is ``rational'' for all $n\ge3$; the
alternative --- eventual transcendence in every dimension, with the plane merely the earliest and only
computable case --- is equally consistent with all current evidence.
\end{question}

The companion papers are recast accordingly: their object, the planar right-triangle group, is the one
dimension in which the amenable quotient of the Coxeter envelope has been carried past its deviation and
found transcendental. Whether that transcendence is a two-dimensional accident or the first instance of a
universal phenomenon is open.

\end{document}
